Tail associations in ecological variables and their impact on extinction risk

### 2020-01-22 MainText.Rmd
new figs added, captions written, and cited from text. just need typo-free versions from Shyamolina

Extinction risk is very commonly estimated for age-structured models 
such as stochastic matrix models [@MorrisDoak2002; @Caswell2000] and 
integral projection models
[@Merowetal2014]. For such models, extinction risk depends, in part, on the long-term
population growth rate, a population parameter which in turn 
depends on the means, variances and covariances of population vital rates such 
as the survival and fecundity rates of age classes. Extinction risks may,
*a priori*, also depend on possible asymmetries of tail association between
population vital rates. For instance, it seems reasonable to imagine
a case in which both adult fecundity and juvenile survival rates in a population 
depend on an environmental variable such as spring rainfall in similar ways, but 
only when that variable takes low values. When rainfall takes
typical or high values, fecundity and survival may be limited instead 
by different, unrelated factors. This could produce left-tail dependence between these 
vital rates, possibly resulting in different extinction risk for the model
compared to a case in which vital rates were symmetrically associated in 
their tails but otherwise statistically similar.

@@

There are also reasons to hypothesize that asymmetry of tail association in 
environmental variables may influence extinction risks for metapopulations.
For instance, assuming for the sake of a simple example that low values of an environmental
variable are "bad" for populations of a given species and high values are "good",
greater left-tail associations between measurements of the variable in different
habitat patches may cause higher metapopulation extinction risk because then
very bad years for component populations would occur at the same time in 
many patches, reducing the potential for rescue effects. Greater right-tail associations
would probably not have the same effect, even if overall correlations between locations
were the same, because in that case it would be very good years for the component 
populations that would occur at the same time in many patches. 
@Ghosh_copula substantiated this intuition by showing that it held true for a 
modeling setup using a spatial version of a very simple non-density-dependent
population growth model (the Lewontin-Cohen model). In essence, if environmental
"catastrophes" (i.e., extremely bad years for a population) are widely 
spatially synchronized it should create much greater metapopulation extinction 
risk than if "bonanzas" (i.e., extremely good years) are widely synchronized, even
if overall (non-tail-specific) levels of environmental synchrony are the same
in both these scenarios. The present study extends tests of this idea to metapopulation
models incorporating density dependence.

@@

>>>>

Extinction risk is very commonly estimated for age-structured models 
such as stochastic matrix models [@MorrisDoak2002; @Caswell2000] and 
integral projection models
[@Merowetal2014]. For such models, extinction risk depends, in part, on the long-term
population growth rate, a population parameter which in turn 
depends on the means, variances and covariances of population vital rates such 
as the survival and fecundity rates of age classes. Extinction risks may,
*a priori*, also depend on possible asymmetries of tail association between
population vital rates. For instance, it seems reasonable to imagine
a case in which both adult fecundity and juvenile survival rates in a population 
depend on an environmental variable such as spring rainfall in similar ways, but 
only when that variable takes low values. When rainfall takes
typical or high values, fecundity and survival may be limited instead 
by different, unrelated factors. This could produce left-tail dependence between these 
vital rates, possibly resulting in different extinction risk for the model
compared to a case in which vital rates are symmetrically associated in 
their tails or right-tail associated but otherwise statistically similar (Fig. \ref{VitalRatePedagogFig}).

@@

There are also reasons to hypothesize that asymmetry of tail association in 
environmental variables may influence extinction risks for metapopulations.
For instance, assuming for the sake of a simple example that low values of an environmental
variable are "bad" for populations of a given species and high values are "good",
greater left-tail associations between measurements of the variable in different
habitat patches may cause higher metapopulation extinction risk because then
very bad years for component populations would occur at the same time in 
many patches, reducing the potential for rescue effects. Greater right-tail associations
would probably not have the same effect, even if overall correlations between locations
were the same, because in that case it would be very good years for the component 
populations that would occur at the same time in many patches. 
@Ghosh_copula substantiated this intuition by showing that it held true for a 
modeling setup using a spatial version of a very simple non-density-dependent
population growth model (the Lewontin-Cohen model). In essence, if environmental
"catastrophes" (i.e., extremely bad years for a population) are widely 
spatially synchronized it should create much greater metapopulation extinction 
risk than if "bonanzas" (i.e., extremely good years) are widely synchronized, even
if overall (non-tail-specific) levels of environmental synchrony are the same
in both these scenarios (Fig. \ref{MetapopPedagogFig}). The present study extends tests 
of this idea to metapopulation models incorporating density dependence.

@@

\begin{figure}
\begin{center}
\includegraphics[width=25cm]{vitalrate_pedagog_fig.pdf}
\caption{Left-tail association between vital rates makes it more likely that low values
of fecundity and survival will happen simultaneously (A), causing both life stages to crash 
simultaneously. This seems likely to increase extinction risk. In contrast, right-tail association
between vital rates instead makes it more likely that high values of fecundity and survival will 
happen simultaneously (B). \label{VitalRatePedagogFig}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=25cm]{metapop_pedagog_fig.pdf}
\caption{Under extreme left-tail association of environments between two locations, catastrophes happen 
simultaneously in both locations (A). Whereas under extreme right-tail association, catastrophes do not happen
simultaneously, even when overall environmental association is the same in both scenarios (B). \label{MetapopPedagogFig}}
\end{center}
\end{figure}
